Combined effects of protein expression variance and correlation on multicomponent systems

### 2020-03-12 Manuscript/Manuscript.Rmd
.

output:
  pdf_document:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    keep_tex: yes
    number_sections: no
    template: manuscript.latex
    includes:
      in_header: header.tex

  word_document: null
  html_document: null

@@

The coordinated expression of proteins is vital for many aspects of cellular function, from maintaining a dynamic steady state to differentiation of cells into specialized types in different tissues. These requirements can range from precise stoichiometric coordination for multisubunit complexes and metabolic pathways, to high levels of noise that can result in heterogeneous responses of identical cell populations to identical stimuli `r citep("10.1073/pnas.1615660114")`. The ability to regulate coordinated expression has been demonstrated to occur via transcription (transcription factors, chromatin regulation), translation (specialized ribosomes, mRNA structure/modifications), and degradation (E3 ligases). These processes, as well as the molecules they target, are all subject to stochastic expression and interactions, which has been shown to lead to differences in cell behavior and decisions in otherwise identical cells. Variability and coordinated expression (correlation) of proteins leads to increased population level control in binary decisions, and a decreased ability to execute analog signaling.

There have been numerous targeted studies of coordinated protein expression and variation that have shown the important impacts of these parameters on cellular function. However, both of these parameters (variation and correlation) act together to determine the total variance of a system.

there is a gap in the literature of direct measurement of the relationship between variation and coordinated protein expression between single cells. To systematically assess these properties of single cells, we carried out a proteomics experiment on exceeding large single cells, Xenopus laevis eggs. An advantage of this approach is that we can get around much of the signal to noise issues that accompany studying single cells due to low sample amounts. Additionally, at this stage of development transcription is restricted so we are able to gain insights into non-transcriptional control of protein expression. We also utilized isobaric tagging in order to measure 25 individual eggs in 5 mass spectrometry runs. In this study, we were able to measure the relative abundance >1000 proteins across single cells in order to better understand the properties of stochastic and coordinated expression of proteins.

With this data set we have been able to, for the first time, measured the relationship between protein expression variance and coordinated protein expression on a proteome scale in single cells. We have observed that certain classes of proteins, including protein complexes and metabolic pathways, are expressed in such a way that increased coordinated expression is balanced by decreased variation. By doing this, cells are able to decrease the total variance of a given complex or pathway, an elegant balancing act that allows for finer control of metabolic throughput and controlling the number of potentially formed complexes though stoichiometric control. Though this kind of coordinated expression leads to an increase in variance in a population of cells, it can reduce variation within a cell.

@@

The specific requirements for coordinated protein expression of pathways or complexes can vary, with some requiring strict stoichiometric regulation with others having much more relaxed requirements. Strict stoichiometry scenarios require coordinated expression of proteins (Figure 1A, top), and others can have uncoordinated expression (Figure 1A bottom). At the single cell level, coordinated expression will result in a high correlation coefficient, whereas un-coordinated expression will result in a low correlation coefficient (Figure 1B). Even when the expression variation (standard deviation or coefficient of variation) and abundance is identical between these two scenarios, the total variance of a pathway or complex can be significantly different due to the variance sum law: $var\left(\sum_{i=1}^n X_{i}\right) = \sum_{i=1}^{n} var(X_{i}) + 2\sum_{i,j:i<j}^{n} cov(X_{i},X_{j})$ (Figure 1C). Our previous work has shown that this effect is important in single cell pathway activation dynamics `r citep("doi.org/10.15252/msb.20177997")`.

In order to study the relationship between protein expression variance and co-expression (correlated expression) of cellular pathways and complexes at the proteome level, we utilized *Xenopus laevis* eggs as a single cell model. Activated *Xenopus laevis* eggs were collected at 5 time points across the first cell cycle (0, 20, 40, 60, and 80 minutes), with 5 eggs at each time point (Fig 1D). Using TMT multiplexing and mass spectrometry, we were able to determine the relative abundance of more than 1300 proteins. Expression of these proteins were largely invariant across the cell cycle, revealing that these highly expressed genes are likely not regulated by cell cycle processes. Additionally, a PCA analysis of these eggs showed no discernible clustering on cell cycle time (Fig S2).

To determine the variance across these proteins, we calculated the variance and coefficient of variation (CV) using all time points collected. This showed a wide range of variation containing multiple distributions (Fig 1E), many of which are consistent with our previous study of variation using targeted mass spectrometry (Fig S2).

Since we were able to measure all of these proteins within single cells, we are able to calculate coordinated expression of protein pairs at single cell resolution. Using Pearson correlation coefficient, we could determine coordinated expression of nearly 2 million protein pairs (Fig 1F-G) The distribution of correlation coefficients fit a normal distribution, with the majority of protein pairs appearing to not show significant co-regulation. However, there appear to be a significant number of protein pairs containing high correlation coefficients, and the heat-map showed a lot of clustering of these highly co-regulated pairs (Fig 1G).

@@

@@

Multicomponent modules such as heteromeric protein complexes and metabolic and signaling pathways have a certain tolerance for noise. Two components of the noise of these systems are the expression variance of the individual proteins, as well as the correlation or covariance of expression between the proteins. This means that if a group of proteins have a positive correlation, then the total variance will be higher than if there was no correlation. This single cell proteomics dataset gave us an excellent opportunity to see how variance and co-expression of proteins are related in low variance and high variance modules. The total variance of a module can be calculated using the variance sum law, where for a group of proteins $X_{1},\dotsc,X_{n}$, the total variance of the module is:
$$\sigma_{{}_{total}}^2 = \sum_{i=1}^{n} \sigma_{X_{i}}^2 + 2\sum_{i,j:i<j}^{n} \rho(X_{i},X_{j})\sigma_{X_i}\sigma_{X_j}$$

To categorize the measured protein expression data into modules, we utilized GO term and KEGG pathway annotations. Since pathways and complexes have varying numbers of proteins, and the number of proteins in a module is highly influential on the total variance sum, we normalized the total variance sum by the number of components (Fig. xA and xB). With this normalized total variance metric, we could separate modules into high and low variance categories (Fig 2A). Since co-expression of proteins (positive correlation between protein expression) increases the total variance, we expected to see the high variance modules to be enriched for high levels of both protein expression variance and co-expression (Fig 1B and 1C, red). However, we were suprised to see that the low variance modules were also enriched for high levels of co-expression of proteins, with some of the mean correlation coefficients of modules as high as those in the high variance category (Fig 1B, orange shaded region).

We wondered how these protein modules dealted with the increased variance load caused by the high numbers of correlated proteins, so we looked at the pairwise relationship between the mean variance of the modules and the mean correlations (Fig S1). The low variance modules uniquely showed a significant negative correlation (R = 0.49) between these two metrics, meaning that for modules that enforce a low level of total variance, co-expression of proteins within that module must also have low expression variance.

@@

## Protein modules display different levels of regulation of noise and co-expression of constituents

We saw that in general, processes xxx were… (Fig. 2D).
(Fig. 3A-D) (Fig. 3E-H)

@@

fig_legend_3 <- c("Density plots of correlation coefficients select protein groups color coded by low (blue) and high (red) normalized protein variance, with the distribution of all measured correlation coefficients colored in grey (rug plot to show the measured correlations coefficents shown on plot). Below each density plot is a box plot.")
```

## Low total variance modules are enriched for metabolic pathways and heteromeric protein complexes

## Enforcement of a trade-off between protein co-expression and noise increases efficiency of metabolic pathways and heteromeric protein complexes

@@

@@

\newpage
# Supplementary material {-}

sup_fig_legend_1 <- c("Pairwise plots of the mean correlation coefficient and mean coefficient of variation of protien groups with linear fit. The low variance protein groups have a significant negative correlation between these two variables, showing that for protein modules that require low variance and high co-expression, showing that these two parameters... R (high, low, med) = (0.057, -0.485, 0.071)")
```

>>>>

output:
  # word_document:
  # html_document:  
  pdf_document:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    keep_tex: yes
    number_sections: no
    template: manuscript.latex
    includes:
      in_header: header.tex

@@

The coordinated expression of proteins is vital for cellular function, from maintaining a dynamic steady state to differentiation of cells into specialized types and tissues. This coordinated regulation of protein abundance is subject to stochastic expression levels, and modules (pathways, complexes, etc.) have varying tolerances of noise for proper function ranging from precise stoichiometric coordination, to high levels of noise that can result in heterogeneous responses of homogeneous cell populations to identical stimuli `r citep("10.1073/pnas.1615660114")`. The ability to regulate coordinated expression has been demonstrated to occur via transcription (transcription factors, chromatin regulation), translation (specialized ribosomes, mRNA structure/modifications), and degradation (E3 ligases), all of which are subject to stochastic expression and interactions. This noise has been shown be utilized by cells for bet hedging strategies as well as tissue size maintenance (more concrete examples to come). Additionally, the combination of variability and coordinated expression (correlation) of proteins can lead to increased population level control in binary decisions, and a decreased ability to execute analog signaling `r citep(c("doi.org/10.15252/msb.20177997", "10.1073/pnas.1615660114"))`.

There have been numerous studies, both targeted and unbiased, of coordinated protein expression or expression variation independently that have shown the important impacts of these parameters on cellular function. However, both of these parameters (variation and correlation) act together to determine the total variance of a system. To systematically assess these properties of single cells at a large scale, we conducted shotgun proteomics analyses on *Xenopus laevis* eggs during their first cell cycle. We have found this model to be ideal at this nascent stage of the single cell proteomics field because of their large size and tractability. An advantage of this approach is that we can get around much of the signal to noise issues that accompany studying single cells caused by low sample protein levels. Additionally, at this stage of development transcription is restricted, allowing for insights into non-transcriptional control of protein expression. By utilizing isobaric tagging and shotgun mass spectrometry we were able to measure >1000 proteins across 25 single cells in 5 mass spectrometry runs.

With this data set we have been able to, for the first time, gain insights into the relationship between protein expression variance, coordinated protein expression, and total variance within modules at a proteomic scale in single cells. We identified classes of proteins that include heteromeric complexes and metabolic pathways that are expressed in such a way that the increase in total variance of the system caused by high coordinated expression of proteins is offset by decreased variation, allowing for modules that rely on stoichiometric expression and low level of variance to maintain a low level of total system variance. By counter balancing coordinated expression (correlation) of proteins in a module with low expression variance, cells are able to decrease the total variance of a given complex or pathway. This elegant balancing act may allow for finer control of metabolic throughput and controlling the number of potentially formed complexes though stoichiometric control.

@@

The specific requirements for coordinated protein expression of pathways or complexes can vary, with some requiring strict stoichiometric regulation and others having much more relaxed requirements. Strict stoichiometry scenarios require coordinated expression of proteins (Figure 1A, top), and others can have uncoordinated expression (Figure 1A bottom). At the single cell level, coordinated expression will result in a high correlation coefficient, whereas un-coordinated expression will result in a low correlation coefficient (Figure 1B). Even when the expression variation (standard deviation or coefficient of variation) and abundance is identical between these two scenarios, the total variance of a pathway or complex can be significantly different due to the variance sum law: $var\left(\sum_{i=1}^n X_{i}\right) = \sum_{i=1}^{n} var(X_{i}) + 2\sum_{i,j:i<j}^{n} cov(X_{i},X_{j})$ (Figure 1C). Our previous work has shown that this effect is important in single cell pathway activation dynamics `r citep("doi.org/10.15252/msb.20177997")`.

In order to study the relationship between protein expression variance and co-expression (correlated expression) of cellular pathways and complexes at the proteome level, we utilized *Xenopus laevis* eggs as a single cell model. Activated *Xenopus laevis* eggs were collected at 5 time points across the first cell cycle (0, 20, 40, 60, and 80 minutes), with 5 eggs at each time point (Fig 1D). Using TMT multiplexing and mass spectrometry, we were able to determine the relative abundance of more than 1300 proteins in single cells. Expression of these proteins across the time course of the cell cycle showed no dynamic patterns, revealing that these highly expressed proteins are likely not regulated by cell cycle processes. Additionally, a PCA analysis of these eggs showed no discernible clustering on cell cycle time (Fig S2).

To determine the expression variability for the measured proteins between single cells, we calculated the variance and coefficient of variation (CV) across all samples. This showed a wide range of variation containing multiple distributions (Fig 1E), many of which are consistent with our previous study of variation using targeted mass spectrometry (Fig S2). An added benefit to measuring these proteins in parallel is that we are able to calculate coordinated expression of protein pairs at single cell resolution. Using the Pearson correlation coefficient, we were able to determine the coordinated expression of nearly 2 million protein pairs (Fig 1F-G). The distribution of correlation coefficients fit a normal distribution centered around 0, with the majority of protein pairs appearing to not show significant co-regulation. However, there appear to be a significant number of protein pairs containing high correlation coefficients, and a clustered heat map shows that many highly co-regulated pairs cluster together (Fig 1G).

Multicomponent modules such as heteromeric protein complexes and metabolic and signaling pathways have a certain tolerance for noise. Two components of the noise of these systems are the expression variance of the individual proteins, as well as the correlation or co-variance of expression between the proteins. This means that if a group of proteins have a positive correlation, then the total variance will be higher than if there was no correlation. This single cell proteomics data set gave us an excellent opportunity to see how variance and co-expression of proteins are related in low variance and high variance modules. The total variance of a module can be calculated using the variance sum law, where for a group of proteins $X_{1},\dotsc,X_{n}$, the total variance of the group is:
$$\sigma_{{}_{total}}^2 = \sum_{i=1}^{n} \sigma_{X_{i}}^2 + 2\sum_{i,j:i<j}^{n} \rho(X_{i},X_{j})\sigma_{X_i}\sigma_{X_j}$$

To categorize the measured protein expression data into modules, we utilized GO term and KEGG pathway annotations. Since pathways and complexes have varying numbers of proteins, and the number of proteins in a module is highly influential on the total variance sum, we normalized the total variance sum by the number of components (Fig. xA and xB) to eliminate this confounding dependency. With the normalized total variance metric, we could separate modules into high and low variance categories (Fig 2A). Since co-expression of proteins (positive correlation between protein expression) increases the total variance, we expected to see the high variance modules to be enriched for high levels of both protein expression variance and co-expression (Fig 1B and 1C, red). However, we were surprised to see that the low variance modules were also enriched for high levels of co-expression of proteins, with some of the mean correlation coefficients of modules as high as those in the high variance category (Fig 1B, orange shaded region).

We wondered how these protein modules dealt with the increased variance load caused by the high numbers of correlated proteins, so we looked at the pairwise relationship between the mean variance of the modules and the mean correlation coefficients (Fig 2D and S1). The low variance modules uniquely showed a significant negative correlation (R = 0.49) between these two metrics, showing that for modules that enforce a low level of total variance, co-expression of proteins within that module is offset by lower expression variance.

## Enforcement of a trade-off between protein co-expression and noise increases efficiency of heteromeric protein complexes and metabolic pathways

Interestingly, we saw the low total variance and high coordinated expression sub population contained many metabolic pathways as well as multisubunit complexes, while the high total variance and high coordinated expression sub-population contained many signalling pathways (Fig. 2D, Fig. 3E-H). Previous studies have shown that protein complex components are produced stoichiometricly `r citep("doi.org/10.1016/j.cels.2018.11.003")` across eukaryotic species and that co-expression of pathway specific enzymes are tightly regulated post-transcriptionally across species `r citep("doi.org/10.1016/j.cell.2018.03.007")`. This led us to wondered what the implications were for the synergy of maintaing low total variance with high levels of co-expression of the members of protein complexes and metabolic pathways.

To test why complexes observed in our data set and others, (i.e. core members of the ribosome (Fig 4A)), coordinate co-expression and expression variation, we imagined an idealized 10 member heteromeric complex (Fig 4B). Coordinated expression could allow for tight control of the total number of assembled complexes. However, this on it's own is not sufficient since high levels of individual protein variance would result in an increased number of partially assembled complexes. In this idealized complex, we titrated the correlation coefficients between all 10 members along with the individual expression variation and saw that increased correlations and decreasing variance had a synergistic effect in allowing for the control of the total number of fully assembled complexes (Fig 4C.). Additionally, there is a large regime with high correlations and low variation that allow for low total variance of the system.

We also observed a number of metabolic pathways that had similar variation and correlation properties as the protein complexes, such as the pentose phosphate pathway and glycolysis. Both of these pathways have a large number of reversible enzymes. For a reversible enzymatic pathway, the concentrations of the enzymes dictate the rate of the reaction, and the concentrations of the reaction products relative to the substrates can dictate the direction of the reaction equilibrium. Therefore, strict balance of the concentration of enzymes using low levels of variance and strict control of the relative abundances of the enzymes is a potentially important step in regulating metabolic pathway output. To test this hypothesis, we constructed an idealized metabolic model (or a physiological model??) and tested the ability of the pathway to maximize output (Fig 5x). (haven't done this, would like feedback / help).

## Discussion

@@

# Figures

@@

@@

@@

fig_legend_3 <- c("Density plots of correlation coefficients select protein groups color coded by low (blue) and high (red) normalized protein variance, with the distribution of all measured correlation coefficients colored in grey (rug plot to show the measured correlations coefficients shown on plot). Below each density plot is a box plot.")
```

@@

# Supplementary material {-}

sup_fig_legend_1 <- c("Pairwise plots of the mean correlation coefficient and mean coefficient of variation of protein groups with linear fit. The low variance protein groups have a significant negative correlation between these two variables, showing that for protein modules that require low variance and high co-expression, showing that these two parameters... R (high, low, med) = (0.057, -0.485, 0.071)")
```

@@
